---
title: "Final Report on Chopped Analysis"
output: github_document
author: "Alana Huitric and Annabelle Platt"
---

# Chopped: Profession or Chance?
<!-- ------------------------------------ -->

## Question
<!-- ------------------------------------ -->

The chefs who participate in Chopped come from a wide variety of cooking backgrounds, from private chefs to home cooks to sous chefs at fancy retaurants. In this analysis, we will investigate the question of whether chefs with more formal training are more likely to win Chopped, or if it is an even playing field. 

## Background
<!-- ------------------------------------ -->

Chopped is a TV show on the Food Network. In each episode four chefs compete in three rounds, the appetizer, entree, and dessert rounds, with one being eliminated, or "chopped", each round. The last chef left standing wins. The catch is that at the beginning of each round, the contestants are given a basket containing ingredients they must incorporate into their dish. These can range from relatively common ingredients, such as bok choy, cream cheese, or strawberries, to much wilder ones, such as gummy fried eggs. The ingredients can also sometimes be challenging to incorporate into the particular dish. A common example of this is all the dessert ingredients being savory. Failure to incorporate a mandatory ingredient into their dish will most likely result in a chef being chopped. 

## Dataset
<!-- ------------------------------------ -->

The data we are using was scraped from the Chopped pages on IMDB. For each episode it contains who the four contestants were, the order in which they were eliminated, information about the contestants' profession, the three judges for that episode, the ingredients for each round, and the episode rating. We also have metadata including season, season episode, and overall series episode. This data can be acquired by running the `get_raw_data` script in this repository. We have also included the resulting data in the `/data` folder. 

```{r}
library(tidyverse)
library(stringr)
library(modelr)
library(stats)

filename = "./data/chopped.csv"
chopped_df <- read_csv(filename)

chopped_df %>% glimpse()
```

## Uncertainty
<!-- ------------------------------------ -->

Because the data we are using is the entire population of Chopped episodes, rather than a sample, there is no uncertainty in our data. 


## Analysis
<!-- ------------------------------------ -->

```{r set-up-theme}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.x = element_text(margin = margin(2, 2, 2, 2), size = 12),
    axis.title.y = element_text(margin = margin(2, 2, 2, 2), size = 12, angle = 90),

    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}
```


Before we could do analysis, there was a lot of data-wrangling to do. This is contained in the next few chunks. 


```{r chef-data-wrangling}
# Make dataframe of only chef names 
df_chef <- chopped_df %>% 
  pivot_longer(
    names_to = "Number",
    values_to = "Chef_Name",
    cols = c(contestant1, contestant2, contestant3, contestant4),
  ) %>% 
  select("Chef_Name","Number","series_episode") 

# Make dataframe of chef info
df_info <- chopped_df %>% 
  pivot_longer(
    names_to = "num",
    values_to = "Info",
    cols = c(contestant1_info, contestant2_info, contestant3_info, contestant4_info),
  ) %>% 
  select("num", "Info")

# Combine chef names and info
df_chef_info_raw <- 
  bind_cols(
    df_chef, df_info
  ) %>% 
  select("Chef_Name", "Info", "series_episode", "num")

df_chef
df_info
df_chef_info_raw
```

```{r convert-place-column}
convert_to_name <- function(x, name) {
  case_when(
    x == TRUE ~ name,
    x == FALSE ~ NA_character_,
  )
}

convert_to_place <- function(x) {
  case_when(
    x == "contestant1_info" ~ "first_chopped",
    x == "contestant2_info" ~ "second_chopped",
    x == "contestant3_info" ~ "third_chopped",
    x == "contestant4_info" ~ "winner"
  )
}

df_chef_info_process <- df_chef_info_raw %>%
  separate(
    col = Info,
    sep = -2,
    into = c("Info","State"),
  ) %>% 
  # Find profession based on string matching
  mutate(
    Private = str_count(Info,"Private") != 0,
    Pastry = str_count(Info,"Pastry")!= 0,
    Catering = str_count(Info, "Catering")!= 0,
    Owner = str_count(Info, "Owner") != 0,
    Executive = str_count(Info, "Executive")!= 0,
    Sous = str_count(Info, "Sous") != 0,
    Restaurateur = str_count(Info, "Restaurateur") != 0,
    Instructor = str_count(Info, "Instructor") != 0,
    IsChef = str_count(Info, "")!= 0
  ) %>% 
  mutate(
    Private = convert_to_name(Private, "Private"),
    Pastry = convert_to_name(Pastry, "Pastry"),
    Catering = convert_to_name(Catering, "Catering"),
    Owner = convert_to_name(Owner, "Owner"),
    Executive = convert_to_name(Executive, "Executive"),
    Sous = convert_to_name(Sous, "Sous"),
    Restaurateur = convert_to_name(Restaurateur, "Restaurateur"),
    Instructor = convert_to_name(Instructor, "Instructor"),
    IsChef = convert_to_name(IsChef, "Chef"),
    place = convert_to_place(num)
  ) %>% 
  unite(
    col = "Profession",
    sep = ", ",
    c(Private, Pastry, Catering, Owner, Executive, Sous, Restaurateur, Instructor, IsChef),
    na.rm = TRUE
  ) %>% 
  select(-Info, -num)
  
df_chef_info_process
```

The following graph shows the raw number of people who entered 

```{r at-first-glance}
df_chef_info_process %>% 
  filter(Profession != "Chef") %>% 
  drop_na() %>% 
  filter(Profession != "") %>% 
  #mutate(Profession = fct_reorder(Profession, ) %>% 
  
  ggplot(aes(Profession, fill = place)) +
  geom_bar() +
  theme_common() + 
  ggtitle("Number of People per Profession by Place") + 
  labs(y = "Count") + 
  coord_flip()
```

Just looking at the raw counts, the distribution of contestant placings is approximately equal within each profession. We will now examine this with proportion instead of raw counts. 

```{r calculate-proportions}
df_prof_prop <- df_chef_info_process %>% 
  group_by(Profession) %>% 
  mutate(
    count_prof = n()
  ) %>% 
  ungroup() %>% 
  group_by(Profession, place) %>% 
  mutate(
    count_prof_place = n(),
    prof_prop = count_prof_place/count_prof,
  ) %>% 
  ungroup()

df_prof_prop %>% glimpse
```
```{r}
df_prop_graph <- df_prof_prop %>%
  filter(prof_prop != 1) %>% 
  mutate(
    Profession = fct_reorder(
      Profession, 
      if_else(place == "winner", prof_prop, -1),
      .fun = max
      )
    ) %>%
  
  ggplot(aes(Profession, prof_prop, color = place, size = count_prof_place)) +
  geom_point() + 
  coord_flip() + 
  ggtitle("Proportion of Contestant Placing by Profession") + 
  labs(y = "Proportion") + 
  geom_hline(yintercept = .25, linetype = 2) + 
  theme_common()

df_prop_graph
```

This graph shows us that in most professions, the proportion of winners falls between about 0.2 and about 0.3, centered around 0.25, which is what we would expect to see profession had no impact on place in the competition. 
There are a few winning proportions that lie outside of this range. 

The size of the points shows us that for proportions outside this range, the number of people in those professions drastically decreased. This suggests they may be a statistical fluke, and we would need more competitors to get a more accurate proportion for that profession. If only a small number of contestants from a particular profession have competed, they will skew the overall proportion of winners in their profession (high if they win, low if they lose). This is why we filtered out professions with perfect winning proportions - because they contained only one or two people. 

Interestingly, we can that as proportions get further from the expected 0.25 value, the count of people in that profession decreases and the points for each place get closer together. This implies that for professions with larger numbers of contestants, the distribution of places is uniform. The proportion for each place in each profession should, as the number of contestants in the profession grows, should approach 0.25 if profession has no impact on place, and this seems to be what we see here. 
 

```{r}
df_prof_prop %>% 
  
  ggplot(aes(count_prof_place, prof_prop)) + 
  geom_violin(draw_quantiles = c(.25, .5, .75)) + 
  coord_flip() +
  facet_wrap(~place) + 
  ggtitle("Density of Profession Proportions in Each Place") + 
  theme(
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.x = element_text(margin = margin(2, 2, 2, 2), size = 12),
    axis.title.y = element_text(margin = margin(2, 2, 2, 2), size = 12, angle = 90),
  ) + 
  labs(y = "Proportion", x = "Count of People in Profession")
```


This graph supports the observations made from the graph above. For each proportion, we can see the combined count of contestants from each profession. Because the data in each case approximately centers around 0.25, we see that the vast majority of professions have an even distribution of places. 

## Conclusions
<!-- ------------------------------------ -->

Based on our analysis, no profession proved more advantageous than any other to winning Chopped, as the majority were fairly evenly distributed. The ones that were not evenly distributed had very few contestants in that profession, and no real conclusion can be made as to the advantage of such a profession, as it has had little impact on the overall winning proportions. As more contestants from those smaller professions compete, we would expect the proportion of all places to approach 0.25. 

## Remaining Questions
<!-- ------------------------------------ -->

What other factors could contribute to winning Chopped, or is it the ultimate even playing field? We did not have any demographic information such as age, sex, or race for the contestants, and location was obfuscated in strings of information. It would also be interesting to analyze the impact of ingredients and favoritism of judges. In particular, the favoritism of judges in combination with demographic information would be interesting. 
